From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-27 Pi/paper/ufin.tex
Resolve comments

We use the type theory of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use
intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for
propositional and set truncation, and set-quotients. All arguments will hold in a Cubical Type
Theory~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}
as well. We review the basics of identity types, homotopy types, and some HITs that we use, and refer the reader to the
book for more details.

@@

A homotopy between functions $f, g : A \to B$, written $f \htpy g$, is given by pointwise equality between them,
$\dfun*{x:A}{f(x) \id_{B} g(x)}$. The identity type for functions is equivalent to homotopies between them ${(f \id_{A
        \to B} g)} \eqv {(f \htpy g)}$, by $\term{funext}$ and $\term{happly}$. An equivalence between types $A \eqv B$ is given
by a pair of functions between them which compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy
  \idfunc_{A}$ (see~\cref{prop:three-chars-equiv} in the appendix), and this is equivalent to the identity type for the
universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, by univalence.

@@

The type $\dsum*{x:A}{P(x)}$ is the collection of all the fibers and is the total space of $P$. The first projection
${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a fibration, that
is, there is a lifting operation (\cref{fig:lift} in the appendix) which lifts paths in the base space to paths in the
total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have:
\[
  \term{lift}(u,p) : (x , u) \id_{\dsum*{x:A}{P(x)}} (y , \transport{P}{p}{u})
\]

% where $\tr{p}{u}$ is shorthand for $$.
\begin{toappendix}
  \begin{figure}
    \begin{center}
      \begin{tikzpicture}[yscale=.5,xscale=2]
        \draw (0,0) arc (-90:170:8ex) node[anchor=south east] {$A$} arc (170:270:8ex);
        \draw (0,6) arc (-90:170:10ex) node[anchor=south east] {$\dsum{x:A}{P(x)}$} arc (170:270:10ex);
        \draw[->] (0,5.8) -- node[auto] {$\fst$} (0,3.2);
        \node[circle,fill,inner sep=1pt,label=left:{$x$}] (b1) at (-.5,1.4) {};
        \node[circle,fill,inner sep=1pt,label=right:{$y$}] (b2) at (.5,1.4) {};
        \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto,swap] {$p$} (b2);
        \node[circle,fill,inner sep=1pt,label=left:{$\pair{x,u}$}] (b1) at (-.5,7.2) {};
        \node[circle,fill,inner sep=1pt,label=right:{$\pair{y,\transport*{P}{p}{u}}$}] (b2) at (.5,7.2) {};
        \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto] {$\term{lift}(u,p)$} (b2);
      \end{tikzpicture}
    \end{center}
    \caption{Lifting operation in $P$}
    \label{fig:lift}
  \end{figure}
\end{toappendix}

@@

Using the groupoid structure of $A$, for any $x, y : A$ and a path $p : x \id_{A} y$, $\term{transport}(p)$ and
$\term{transport}(\inv{p})$ form an equivalence.
\[
  \tptEqv{P} : (x \id_{A} y) \to (P(x) \eqv P(y))
\]

>>>>

We use the type theory of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use
intensional Martin-L\"{o}f Type Theory, with a hierarchy of (univalent) universes $\UU_0 : \UU_1 : \ldots$ (though we
will just write $\UU$ since we mostly use one universe), and a few Higher Inductive Types (HITs) for truncations and
quotients. We write $\dfun*{x:A}{B(x)}$ or simply $(x:A) \to B(x)$ for dependent function types, and $\dsum*{x:A}{B(x)}$
or simply $(x:A) \times B(x)$ for dependent sum types.

All arguments will hold in a Cubical Type
Theory~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}
as well. We review the basics of identity types, homotopy types, and some HITs that we use, and refer the reader to the
book for more details.

@@

A homotopy between functions $f, g : A \to B$, written $f \htpy g$, is given by pointwise equality between them,
$\dfun*{x:A}{f(x) \id_{B} g(x)}$. The identity type for functions is equivalent to homotopies between them ${(f \id_{A
        \to B} g)} \eqv {(f \htpy g)}$, by $\term{funext}$ and $\term{happly}$. An equivalence between types $A \eqv B$ is given
by a pair of functions between them which compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy
  \idfunc_{A}$ (also see~\cref{prop:three-chars-equiv} in the appendix), and this is equivalent to the identity type for
the universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, which is the univalence principle.

@@

The type $\dsum*{x:A}{P(x)}$ is the collection of all the fibers and is the total space of $P$. The first projection
${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a fibration, that
is, there is a lifting operation (\cref{fig:lift} in the appendix) which lifts paths in the base space to paths in the
total space.

\begin{toappendix}
  Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have:
  \[
    \term{lift}(u,p) : (x , u) \id_{\dsum*{x:A}{P(x)}} (y , \transport{P}{p}{u})
  \]

  % where $\tr{p}{u}$ is shorthand for $$.

  \begin{figure}
    \begin{center}
      \begin{tikzpicture}[yscale=.5,xscale=2]
        \draw (0,0) arc (-90:170:8ex) node[anchor=south east] {$A$} arc (170:270:8ex);
        \draw (0,6) arc (-90:170:10ex) node[anchor=south east] {$\dsum{x:A}{P(x)}$} arc (170:270:10ex);
        \draw[->] (0,5.8) -- node[auto] {$\fst$} (0,3.2);
        \node[circle,fill,inner sep=1pt,label=left:{$x$}] (b1) at (-.5,1.4) {};
        \node[circle,fill,inner sep=1pt,label=right:{$y$}] (b2) at (.5,1.4) {};
        \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto,swap] {$p$} (b2);
        \node[circle,fill,inner sep=1pt,label=left:{$\pair{x,u}$}] (b1) at (-.5,7.2) {};
        \node[circle,fill,inner sep=1pt,label=right:{$\pair{y,\transport*{P}{p}{u}}$}] (b2) at (.5,7.2) {};
        \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto] {$\term{lift}(u,p)$} (b2);
      \end{tikzpicture}
    \end{center}
    \caption{Lifting operation in $P$}
    \label{fig:lift}
  \end{figure}
\end{toappendix}

@@

Using the groupoid structure of $A$, for any $x, y : A$ and a path $p : x \id_{A} y$, $\term{transport}(p)$ and
$\term{transport}(\inv{p})$ form an equivalence, lifting paths in the base space to equivalences between fibers.
\[
  \tptEqv{P} : (x \id_{A} y) \to (P(x) \eqv P(y))
\]
